---
title: "class1_R_2025"
author: "Babatunde Oyedele_101385"
date: "2025-10-16"
output: html_document
---

# Important definitions in data science
 Statistics
 Modelling
 Computation
 Calculus
# Today’s class,we will study
 Installation of R
 Installation of R studio
 Calculator in r
 Data importation in R
 Data description using numerical measures and graphs.
# Data structure (Vectors, Factors, Lists, Data frame, matrix and arrays)
# Vector: The foundational R data structure
# Numeric
```{r}
 x <- c(0.5, 0.6)
 age<-c(20,35,32,29)
 summary(age)
 ## Min. 1st Qu. Median Mean 3rd Qu. Max.
 ## 20.00 26.75 30.50 29.00 32.75 35.00
 age<-c(35,24,18,24)
 mean(age)
 ## [1] 25.25
```
# Logical
```{r}
 x <- c(TRUE, FALSE)
 x <- c(T, F)
```

# character
```{r}
x<-c("a","b","c")
class<-c("M","F","F","M")
```

# Integer
```{r}
x<-9:29
```

# Complex
```{r}
 x<-c(1+0i,2+4i)
```

# Matrixandoperationwithinmatrices
```{r}
 x<-matrix(1:6,nrow=2,ncol=3)
 x
 ## [,1][,2][,3]
 ##[1,] 1 3 5
 ##[2,] 2 4 6
 (y<-matrix(1:6,nrow=3,ncol=2))
 ## [,1][,2]
 ##[1,] 1 4
 ##[2,] 2 5
 ##[3,] 3 6
 x%*%y
 ## [,1][,2]
 ##[1,] 22 49
 ##[2,] 28 64
```

# List
```{r}
 
 x<-list(1,"a",TRUE,1+4i)
 x
 ##[[1]]
 ##[1]1
 ##
 ##[[2]]
 ##[1]"a"
 ##
 ##[[3]]
 ##[1]TRUE
 ##
 ##[[4]]
 ##[1]1+4i
```

# Factor
```{r}
 x<-factor(c("yes","yes","no","yes","no"))
```

# CreateavectorwithNAsinit
```{r}
 x<-c(1,2,NA,10,3)
```

Return a logical vector indicating which elements are NA
```{r}
is.na(x)
 ##[1]FALSEFALSE TRUEFALSEFALSE
 x<-c(1,2,4,"NA",5)
 bad<-is.na(x)
 print(bad)
 ##[1]FALSEFALSEFALSEFALSEFALSE
 x[!bad]
 ##[1]"1" "2" "4" "NA""5"
```

What if there are multiple R objects and you want to take the subset with no missing values in any of those objects?
```{r}
 x<-c(1,2,NA,4,NA,5)
 y<-c("a","b",NA,"d",NA,"f")
 good<-complete.cases(x,y)
 good
 ##[1] TRUE TRUEFALSE TRUEFALSE TRUE
```

# Coercion
 
If character is present,in a vector,R convert everything in the vector to character
strings.
If a vector only contains logical and numbers,R will convert the logical to numbers, Every true becomes a 1,and every FALSE becomes 0
```{r}

 sum(c(TRUE,TRUE,FALSE,FALSE,FALSE))
 ##[1]2
```

# Createdata

Create data Using dataframe 
Dataframe is more general than matrix How???
Because dataframe can contain different modes of data (Numeric, character and so on) Similar to what you can see in SPSS,SAS,…
 
 ###Let’s create a data frame
```{r}
studentID<-c(1,2,3,4,5)
 math_score<-c(12,17,10,9,NA)
 gender<-c("M","F","M","M","F")
 it_score<-c(13,18,11,10,19)
 scoredata<-data.frame(studentID,gender,math_score,it_score)
 scoredata
```

# Createdatafromkeyboard
Steps1.Create dataframe(ormatrix) with variable names
steps2.Invoke the text editor in the data objected created at first step
 
```{r}
data_class2<-data.frame(height=numeric(0),weight=numeric(0),bmi=numeric
 (0))
data_class2<-edit(data_class2)
```

# Data Importation
 R has some features that can allow to import data from different sources(It can be
 text file,spreadsheet,or database)
 1. Data from excel
 2. Data statistical packages(SAS,SPSS,Stata)
 3. Data from Textfiles(ASCII,XML,Webscraping)
 4. Data from database management systems(SQL,MySQL,Oracle,Access)

# HW1:Import data from Statistical package and from Database management systems
```{r}
data_class<-read.table("C:\\Users\\Pacy\\OneDrive\\Desktop\\Bigdatacourse\\class_data.txt")

variable.names(data_class)
##[1]"HEIGHT""WEIGHT"
 
head(data_class)

 
tail(data_class)

data_class[10:20,]

summary(data_class$WEIGHT)


length(data_class$WEIGHT)

data_class[,-1]
```

Incase you want to use data set built in R

```{r}
data()#listofdatasetscurrentlyavailable
data("airquality")
variable.names(airquality)
 ##[1]"Ozone" "Solar.R""Wind" "Temp" "Month" "Day"
str(airquality)
 ##'data.frame': 153obs.of 6variables:
 ## $Ozone :int 41361218NA2823198NA...
 ## $Solar.R:int 190118149313NANA2999919194...
 ## $Wind :num 7.4812.611.514.314.98.613.820.18.6...
 ## $Temp :int 67727462566665596169...
 ## $Month :int 5555555555...
 ## $Day :int 12345678910...
```


Explorationdataanalysis

Examinenumericalvariableusingcommonsummary
```{r}
data_class
summary(data_class)
str(data_class)
```

# SummaryStatistics
```{r}
 summary(data_class)
 ## HEIGHT WEIGHT
 ## Min. :150.0 Min. :42.0
 ## 1stQu.:156.2 1stQu.:48.0
 ## Median:159.5 Median:52.0
 ## Mean :159.4 Mean :52.4
 ## 3rdQu.:162.0 3rdQu.:56.0
 ## Max. :170.0 Max. :65.0
 apply(data_class,2,mean)
 ## HEIGHT WEIGHT
 ##159.38095 52.40476
 apply(data_class,2,sd)
 ## HEIGHT WEIGHT
 ##4.2767235.806050
 c(mean(data_class$HEIGHT),sd(data_class$HEIGHT))
 ##[1]159.380952 4.276723
 c(mean(data_class$WEIGHT),sd(data_class$WEIGHT))
 ##[1]52.40476 5.80605
 c(Mean=mean(data_class$HEIGHT),SD=sd(data_class$HEIGHT))
 ## Mean SD
 ##159.380952 4.276723
 c(Mean=mean(data_class$WEIGHT),SD=sd(data_class$WEIGHT))
```

#Variationandcovariationusingbothnumericalandgraphs
 
## Relationshipbetweenacategoricalandacontinuousvariable


```{r}

 data("iris")
 head(iris)
 ## Sepal.LengthSepal.WidthPetal.LengthPetal.WidthSpecies
 ##1 5.1 3.5 1.4 0.2 setosa
 ##2 4.9 3.0 1.4 0.2 setosa
 ##3 4.7 3.2 1.3 0.2 setosa
 ##4 4.6 3.1 1.5 0.2 setosa
 ##5 5.0 3.6 1.4 0.2 setosa
 ##6 5.4 3.9 1.7 0.4 setosa
 mean(iris$Petal.Length)
 ##[1]3.758
 mean(iris$Petal.Length[iris$Species=="setosa"])
 ##[1]1.462
 mean(iris$Petal.Length[iris$Species=="versicolor"])
 ##[1]4.26
 mean(iris$Petal.Length[iris$Species=="virginica"])
 ##[1]5.552
 ##shortcut
 by(iris$Petal.Length,iris$Species,mean)
 ##iris$Species:setosa
 ##[1]1.462
 ##------------------------------------------------------------
 ##iris$Species:versicolor
 ##[1]4.26
 ##------------------------------------------------------------
 ##iris$Species:virginica
 ##[1]5.552
 by(iris$Petal.Length,iris$Species,sd)
 
by(iris$Petal.Length,iris$Species,summary)
```

# Visualize the differences in continuous variables between categories using Box-and-whisker plot
```{r}
 boxplot(iris$Sepal.Length~iris$Species,data=iris, main="Comparison")
```

# Relationship between two categorical variables
```{r}
 ucba<-data.frame(UCBAdmissions)
 head(ucba)
 ##
 Admit Gender Dept Freq
 ## 1 Admitted Male A 512
 ## 2 Rejected Male A 313
 ## 3 Admitted Female
 A 89
 ## 4 Rejected Female
 A 19
##5Admitted Male B 353
 ##6Rejected Male B 207
 cross<-xtabs(Freq~Gender+Admit,data=ucba)
 (cross<-xtabs(Freq~Gender+Admit,data=ucba))
 ## Admit
 ##Gender AdmittedRejected
 ## Male 1198 1493
 ## Female 557 1278
 ##IstheregenderbiasinUCBgraduateadmissionprocess?
 prop.table(cross,2)
 ## Admit
 ##Gender Admitted Rejected
 ## Male 0.68262110.5387947
 ## Female0.31737890.4612053
```


#  Simpson’sparadox
Phenomenon,whereatrendthatappearsincombinedgroupsofdatadisappearsor reverseswhenbrokendownintogroups
```{r}
cross2<-xtabs(Freq~Gender+Admit,data=ucba[ucba$Dept=="A",])
prop.table(cross2,1)
```

#  BarChart
```{r}

 dat<-read.table(text="ProdAProdBProdCProdD
 1110506070
 2120508065",header=TRUE)

 barplot(as.matrix(dat),beside=FALSE,col=c("Red","green"))

```

```{r}
#barplot(as.matrix(dat),beside=TRUE,col=c("gold3","red"))
 dat <- read.table(text = "A B C D E F G
 1 10 80 30 90 70 60 90
 2 20 50 70 50 40 10 40
 3 60 80 80 60 60 30 160
 4 20 40 70 80 20 10 70", header = TRUE)
 barplot(as.matrix(dat))
```

# Scatter Plot
```{r}
 plot(WEIGHT~HEIGHT,data=data_class)
```

```{r}
plot(data_class$HEIGHT,data_class$WEIGHT)
```

```{r}
plot(data_class[,1],data_class[,2])
```

```{r}
cor(data_class$HEIGHT,data_class$WEIGHT)
 ## [1] 0.2853684
```

# Linear Models
```{r}

 plot(WEIGHT~HEIGHT,data=data_class)
 abline(lm(WEIGHT~HEIGHT,data=data_class)$coefficient)
```


```{r}
data_lm<-lm(WEIGHT~HEIGHT,data=data_class)
 summary(data_lm)
 
hist(data_class$HEIGHT)
```

```{r}
hist(data_class$WEIGHT,col = "blue")
```

# Linear with other line
```{r}
 plot(WEIGHT~HEIGHT,data=data_class)
 abline(lm(WEIGHT~HEIGHT,data=data_class)$coefficient)
 lines(lowess(data_class$HEIGHT,data_class$WEIGHT),col="blue")
```

```{r}
cor(data_class$HEIGHT,data_class$WEIGHT)
 ## [1] 0.2853684
 model1<-lm(WEIGHT~HEIGHT,data=data_class)
```

# Data management
# Create another Variable

```{r}
 attach(data_class)
 (BMI<-WEIGHT/(HEIGHT/100)^2)
 ## [1] 19.28938 20.39542 16.82423 22.49135 23.43750 21.36752 22.10029
 18.42653
 ## [9] 18.02596 18.28678 23.43750 17.48179 22.34778 19.97919 19.47341
 19.47341
 ## [17] 20.70312 20.83000 19.92187 20.70312 19.04432 23.61275 28.00000
 20.06095
 ## [25] 21.71925 18.21832 20.06095 22.47659 18.19548 18.81892 24.12879
 21.48437
 ## [33] 18.02596 23.14050 24.65483 21.07720 21.64412 17.47472 19.57168
 19.48696
 ## [41] 19.47341 21.56454
 (BMI<-round(WEIGHT/(HEIGHT/100)^2,digit=2))
 
 head(cbind(data_class,BMI),n=5)
 
 new_data_class<-cbind(data_class,BMI)
 tail(cbind(data_class,BMI),n=10)
```

# SummaryofBMI
```{r}
summary(BMI)
 ## Min.1stQu. Median Mean3rdQu. Max.
 ## 16.82 19.10 20.23 20.64 22.00 28.00
stem(BMI,scale=2)
```

